---
title: "PCA from Scratch"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 5.5)
```

PCA finds a new set of axes for your data. The first points along the direction of
greatest variation, the second along the greatest remaining variation at right angles
to it, and so on. Nothing is discarded and nothing is transformed — only the coordinate
system changes.

This notebook builds that from the covariance matrix on two correlated variables, where
the geometry is visible, then checks the result against `prcomp()`. Everything here
scales directly to thousands of genes.

```{r}
library(ellipse)
```

```{r echo=FALSE}
dir.create("../figures", showWarnings = FALSE)
dir.create("../results", showWarnings = FALSE)
```

---

## A toy dataset

Two variables that move together, the way two co-expressed genes would.

```{r}
set.seed(42)
n  <- 200
x1 <- rnorm(n, 0, 3)
x2 <- 0.8 * x1 + rnorm(n, 0, 1.2)
X  <- cbind(x1, x2)

plot(X, pch = 19, col = "#1B6B70AA", asp = 1,
     xlab = "Gene 1", ylab = "Gene 2",
     main = "Two correlated features")
abline(h = 0, v = 0, col = "grey80", lty = 3)
```

The cloud is a tilted ellipse, and that tilt is the correlation. Knowing Gene 1 tells
you something about Gene 2, which means the two axes carry redundant information.

---

## Centre, then compute the covariance matrix

```{r}
Xc <- scale(X, center = TRUE, scale = FALSE)   # centre only
C  <- cov(Xc)
round(C, 3)
```

The diagonal is each variable's own variance. The off-diagonal is the covariance between
them — non-zero, which is that tilt expressed numerically. PCA looks for a rotation in
which it becomes zero.

---

## Eigendecomposition

The eigenvectors of the covariance matrix are the new axes; the eigenvalues are how much
variance lies along each.

```{r}
e <- eigen(C)

round(e$values, 3)
round(e$vectors, 3)

cat("variance explained:", round(100 * e$values / sum(e$values), 1), "%\n")
```

The first eigenvalue dwarfs the second. Two correlated variables really carry close to
one variable's worth of information, and PCA has found it.

---

## The components, drawn on the data

The 95% ellipse *is* the covariance matrix drawn as a shape. The principal components
are its major and minor axes.

```{r}
s <- 3 * sqrt(e$values)     # arrow length proportional to SD along each axis

draw_components <- function() {
  plot(Xc, pch = 19, col = "#1B6B70AA", asp = 1,
       xlab = "Gene 1 (centred)", ylab = "Gene 2 (centred)",
       main = "Principal components")
  lines(ellipse(C, level = 0.95), col = "grey60", lwd = 1.5, lty = 2)
  arrows(0, 0, s[1] * e$vectors[1, 1], s[1] * e$vectors[2, 1],
         col = "#E07A5F", lwd = 3, length = 0.12)
  arrows(0, 0, s[2] * e$vectors[1, 2], s[2] * e$vectors[2, 2],
         col = "#2A4B7C", lwd = 3, length = 0.12)
  legend("topleft", c("PC1", "PC2", "95% ellipse"),
         col = c("#E07A5F", "#2A4B7C", "grey60"),
         lwd = c(3, 3, 1.5), lty = c(1, 1, 2), bty = "n")
}

draw_components()
```

```{r echo=FALSE, results='hide'}
svg("../figures/pca_components_toy.svg", width = 6, height = 5.5)
draw_components()
dev.off()
```

PC1 runs along the long axis of the ellipse. PC2 is perpendicular to it and much
shorter, because there is less variation in that direction.

---

## The rotation

Projecting onto the new axes is a single matrix multiplication.

```{r}
Z <- Xc %*% e$vectors
```

Each row of `Z` is the same sample expressed in the new principal-component coordinate
system — same measurement, different axes.

```{r fig.width=10, fig.height=5}
draw_rotation <- function() {
  par(mfrow = c(1, 2))
  plot(Xc, pch = 19, col = "#1B6B70AA", asp = 1,
       main = "Original space", xlab = "Gene 1", ylab = "Gene 2")
  plot(Z,  pch = 19, col = "#E07A5FAA", asp = 1,
       main = "PC space", xlab = "PC1", ylab = "PC2")
  par(mfrow = c(1, 1))
}

draw_rotation()
```

```{r echo=FALSE, results='hide'}
svg("../figures/pca_rotation_toy.svg", width = 10, height = 5)
draw_rotation()
dev.off()
```

```{r}
round(cov(Z), 6)
```

The off-diagonal is now zero. That is the whole point: PCA does not change the data, it
changes the coordinate system so the axes are uncorrelated. The two panels above show
the same 200 points viewed from a different angle.

---

## Checking against `prcomp()`

```{r}
p <- prcomp(X, center = TRUE, scale. = FALSE)

round(p$sdev^2, 3)       # matches the eigenvalues
round(p$rotation, 3)     # matches the eigenvectors
```

Identical. `prcomp()` does the same operation via singular value decomposition, which is
numerically more stable but mathematically equivalent.

One quirk worth knowing: eigenvectors are defined only up to sign, so `prcomp()` may
point an axis the opposite way to a manual calculation. A PCA plot that looks mirrored
is usually this, not a bug.

---

## Why scaling matters

Suppose Gene 2 had been measured in different units — the same biology, a factor of 100
larger.

```{r}
X2 <- cbind(gene1 = x1, gene2 = x2 * 100)

p_unscaled <- prcomp(X2, scale. = FALSE)
p_scaled   <- prcomp(X2, scale. = TRUE)

cat("unscaled — variance explained:",
    round(100 * p_unscaled$sdev^2 / sum(p_unscaled$sdev^2), 1), "%\n")
round(p_unscaled$rotation, 3)
```

```{r}
cat("scaled — variance explained:",
    round(100 * p_scaled$sdev^2 / sum(p_scaled$sdev^2), 1), "%\n")
round(p_scaled$rotation, 3)
```

```{r}
scaling <- data.frame(
  version = c("unscaled", "scaled"),
  PC1_pct = c(round(100 * p_unscaled$sdev[1]^2 / sum(p_unscaled$sdev^2), 1),
              round(100 * p_scaled$sdev[1]^2   / sum(p_scaled$sdev^2), 1)),
  gene1_PC1_loading = c(round(p_unscaled$rotation[1, 1], 3),
                        round(p_scaled$rotation[1, 1], 3)),
  gene2_PC1_loading = c(round(p_unscaled$rotation[2, 1], 3),
                        round(p_scaled$rotation[2, 1], 3))
)
scaling
```

```{r echo=FALSE}
write.csv(scaling, "../results/scaling_effect_toy.csv", row.names = FALSE)
```

Unscaled, PC1 loads -1.00 on `gene2` and -0.01 on `gene1`, and reports 100% of the
variance explained. Nothing about the biology changed — only the units. PCA maximises
variance, and variance depends on the scale a variable is measured on.

Scaled, both genes load equally (-0.707 each) and the original structure comes back.

---

## When to scale

- **Different units** — expression alongside age, BMI, or clinical scores. Scale, or the
  variable with the largest numeric range dominates.
- **Bulk RNA-seq after VST or rlog** — usually don't. Genes are already on a comparable
  log scale, and scaling inflates the noise in lowly-expressed ones.
- **scRNA-seq** — the usual pipeline scales after selecting highly variable genes, which
  limits the damage. Scaling all genes amplifies noise in the many near-zero ones.

The transformation you apply before PCA is not a formality. It determines what PC1
turns out to be.

---

## What carries over

Everything here holds with 20,000 genes instead of two. The covariance matrix is
20,000 × 20,000 rather than 2 × 2, and you keep the first handful of components instead
of looking at both — but it is the same rotation, chosen the same way.

The next notebook applies this to a real bulk RNA-seq experiment, where PC1 turns out to
be a specific, nameable biological programme — provided the data is transformed
correctly first.
