---
title: "PCA on Bulk RNA-seq"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4.5)
```

Bulk RNA-seq is where most people meet PCA. Almost every paper has a PCA plot near the
front, usually captioned "samples cluster by condition" and left at that.

This notebook works through what PC1 actually is on a real experiment, what happens if
you skip the transformation step, and how to decide how many components to keep.

**Data:** `airway` — human airway smooth muscle cells from four donors, each with and
without dexamethasone treatment. Eight samples, paired design.
Himes et al. (2014), *PLoS One* 9:e99625.

```{r}
library(DESeq2)
library(airway)

pal <- c("#1B6B70", "#E07A5F")

data(airway)
airway
```

```{r echo=FALSE}
dir.create("../figures", showWarnings = FALSE)
dir.create("../results", showWarnings = FALSE)
```

---

## The standard plot

`vst()` applies a variance-stabilising transformation. Raw count variance grows with the
mean — a gene at 10,000 counts varies far more in absolute terms than one at 10, purely
from Poisson sampling. VST removes that dependence so PCA sees biological variation
rather than counting noise.

```{r}
dds <- DESeqDataSet(airway, design = ~ dex)
vsd <- vst(dds, blind = TRUE)

plotPCA(vsd, intgroup = "dex")
```

PC1 (41%) separates treated from untreated cleanly. The experiment worked.

```{r}
plotPCA(vsd, intgroup = "cell")
```

The same points coloured by donor. PC2 (26%) is the cell line — each donor appears once
on each side of PC1, because every donor received both treatments.

That structure is worth acting on. The paired design belongs in the model:

```{r eval=FALSE}
dds2 <- DESeqDataSet(airway, design = ~ cell + dex)
```

Fitting `~ dex` alone leaves donor differences in the residual variance and costs power.

---

## What happens without VST

The same PCA on plain log2 counts.

```{r}
rld_raw <- log2(counts(dds) + 1)
pca_raw <- prcomp(t(rld_raw[rowSums(counts(dds)) > 10, ]))

plot(pca_raw$x[, 1], pca_raw$x[, 2],
     col = pal[as.integer(colData(dds)$dex)], pch = 19, cex = 1.4,
     xlab = "PC1", ylab = "PC2", main = "Without VST")
```

The groups still separate — but along a diagonal rather than along PC1, and the points
fall on a curve. That arc is the signature of unstabilised variance.

```{r}
pv_raw <- pca_raw$sdev^2 / sum(pca_raw$sdev^2)
round(pv_raw[1:4] * 100, 1)

pca_vst <- prcomp(t(assay(vsd)[order(rowVars(assay(vsd)), decreasing = TRUE)[1:500], ]))
pv_vst  <- pca_vst$sdev^2 / sum(pca_vst$sdev^2)
round(pv_vst[1:4] * 100, 1)
```

---

## What PC1 really is without VST

The obvious suspect is sequencing depth. Testing it directly:

```{r}
libsize <- colSums(counts(dds))

cor(pca_raw$x[, 1], libsize)
cor(pca_vst$x[, 1], libsize)
```

Without VST, PC1 correlates **-0.89** with library size. After VST, **-0.17**.

So the first plot was not showing treatment. It was showing how deeply each sample was
sequenced, with treatment happening to correlate with it across eight samples. VST
removes that and PC1 becomes biology.

```{r}
pca_summary <- data.frame(
  transform       = c("raw log2(counts+1)", "VST"),
  PC1_pct         = c(round(pv_raw[1] * 100, 1), round(pv_vst[1] * 100, 1)),
  PC2_pct         = c(round(pv_raw[2] * 100, 1), round(pv_vst[2] * 100, 1)),
  PC1_cor_libsize = c(round(cor(pca_raw$x[, 1], libsize), 3),
                      round(cor(pca_vst$x[, 1], libsize), 3))
)
pca_summary
```

```{r echo=FALSE}
write.csv(pca_summary, "../results/vst_effect_on_pca.csv", row.names = FALSE)
```

With eight samples a correlation of -0.89 has wide uncertainty. The direction is not in
doubt and the mechanism is well established, but treat the exact figure loosely.

---

## Naming PC1

A component is only interpretable if you can say which genes make it up. The loadings
are the coefficients that define the axis.

```{r}
loadings <- pca_vst$rotation[, 1]
top      <- sort(abs(loadings), decreasing = TRUE)[1:20]
genes    <- rowData(airway)$symbol[match(names(top), rownames(assay(vsd)))]

pc1_genes <- data.frame(gene = genes, loading = round(loadings[names(top)], 3))
pc1_genes
```

```{r echo=FALSE}
write.csv(pc1_genes, "../results/pc1_loadings_airway.csv", row.names = FALSE)
```

FKBP5, TSC22D3 (GILZ), KLF15, ZBTB16, DUSP1 — canonical glucocorticoid receptor targets.
PC1 is the glucocorticoid response programme, and you can name the genes rather than
gesturing at "condition".

VCAM1 loads negatively while the rest are positive: an inflammatory adhesion molecule
going down as the anti-inflammatory programme goes up. The sign of a loading carries
information.

---

## Biplot

Samples and gene loadings on the same axes. Arrow direction shows which way a gene pulls
samples; length shows how strongly.

```{r fig.width=8, fig.height=7}
top10   <- names(sort(abs(loadings), decreasing = TRUE)[1:10])
sym10   <- rowData(airway)$symbol[match(top10, rownames(assay(vsd)))]
arrows_ <- pca_vst$rotation[top10, 1:2]

sc <- 0.7 * max(abs(pca_vst$x[, 1:2])) / max(abs(arrows_))

lab_x <- arrows_[, 1] * sc * 1.20
lab_y <- arrows_[, 2] * sc * 1.20 + seq(-2, 2, length.out = length(sym10))

draw_biplot <- function() {
  par(mar = c(4.5, 4.5, 3, 1))
  plot(pca_vst$x[, 1], pca_vst$x[, 2],
       col = pal[as.integer(colData(dds)$dex)], pch = 19, cex = 1.4,
       xlab = paste0("PC1 (", round(pv_vst[1] * 100, 1), "%)"),
       ylab = paste0("PC2 (", round(pv_vst[2] * 100, 1), "%)"),
       main = "Biplot: samples and top PC1 genes",
       xlim = range(pca_vst$x[, 1]) * 1.3,
       ylim = range(pca_vst$x[, 2]) * 1.3)
  abline(h = 0, v = 0, col = "grey85", lty = 3)
  arrows(0, 0, arrows_[, 1] * sc, arrows_[, 2] * sc,
         col = "grey40", length = 0.08, lwd = 1.5)
  text(lab_x, lab_y, labels = sym10, cex = 0.65, col = "grey20")
  legend("topleft", levels(colData(dds)$dex), col = pal, pch = 19,
         bty = "n", cex = 0.8)
}

draw_biplot()
```

```{r echo=FALSE, results='hide'}
svg("../figures/biplot_airway.svg", width = 8, height = 7)
draw_biplot()
dev.off()
```

Every arrow except VCAM1 points right, towards the treated samples. That is the
glucocorticoid programme turning on. VCAM1 points left, towards untreated — the
inflammatory gene being suppressed.

A loadings table gives you the same numbers, but the biplot shows direction and
magnitude in the same space as the samples.

---

## Scree plot

```{r}
barplot(pv_vst[1:7] * 100, names.arg = paste0("PC", 1:7),
        ylab = "% variance", main = "Scree plot (VST)",
        col = "#1B6B70", border = NA)
```

```{r echo=FALSE, results='hide'}
png("../figures/scree_airway.png", width = 1800, height = 1400, res = 300)
barplot(pv_vst[1:7] * 100, names.arg = paste0("PC", 1:7),
        ylab = "% variance", main = "Scree plot (VST)",
        col = "#1B6B70", border = NA)
dev.off()
```

A clear break after PC4, then almost nothing. With eight samples there are only seven
non-zero components, so this scree is short by construction.

---

## How many PCs, and how much is lost

An alternative to eyeballing the elbow: pick the smallest number of components reaching
some cumulative variance, then measure how much information the truncation discards by
reconstructing the data and comparing.

```{r}
X  <- t(assay(vsd)[order(rowVars(assay(vsd)), decreasing = TRUE)[1:500], ])
Xc <- scale(X, center = TRUE, scale = FALSE)

cum <- cumsum(pv_vst)
n90 <- which(cum >= 0.90)[1]
cat("PCs needed for 90% of variance:", n90, "of", ncol(pca_vst$x), "\n")

recon_err <- sapply(1:ncol(pca_vst$x), function(k) {
  Xhat <- pca_vst$x[, 1:k, drop = FALSE] %*% t(pca_vst$rotation[, 1:k, drop = FALSE])
  mean((Xc - Xhat)^2)
})

recon <- data.frame(
  n_pcs          = seq_along(recon_err),
  cumulative_pct = round(cum * 100, 1),
  mse            = round(recon_err, 4)
)
recon
```

```{r echo=FALSE}
write.csv(recon, "../results/reconstruction_error_airway.csv", row.names = FALSE)
```

```{r}
par(mar = c(4.5, 4.5, 3, 1))
plot(recon$n_pcs, recon$mse, type = "b", pch = 19, col = "#1B6B70",
     xlab = "Number of PCs retained", ylab = "Reconstruction MSE",
     main = "Information lost by truncating")
abline(v = n90, lty = 2, col = "#E07A5F")
text(n90 + 0.2, max(recon$mse) * 0.8, paste0("90% variance\nat ", n90, " PCs"),
     col = "#E07A5F", cex = 0.7, pos = 4)
```

```{r echo=FALSE, results='hide'}
png("../figures/reconstruction_airway.png", width = 1800, height = 1400, res = 300)
par(mar = c(4.5, 4.5, 3, 1))
plot(recon$n_pcs, recon$mse, type = "b", pch = 19, col = "#1B6B70",
     xlab = "Number of PCs retained", ylab = "Reconstruction MSE",
     main = "Information lost by truncating")
abline(v = n90, lty = 2, col = "#E07A5F")
text(n90 + 0.2, max(recon$mse) * 0.8, paste0("90% variance\nat ", n90, " PCs"),
     col = "#E07A5F", cex = 0.7, pos = 4)
dev.off()
```

Four components reach 96.8% and the reconstruction error drops from 0.754 to 0.041. The
curve flattens exactly where the scree plot breaks.

This works cleanly on bulk data. It is a poor guide for single-cell: in notebook 03,
fifty PCs reach only 14% of the variance, and demanding 90% would mean keeping hundreds.
What you discard there is mostly dropout noise, so reconstruction error would be
measuring how well you reproduce noise.

---

## What to take from this

- **Transform first.** Without VST, PC1 was sequencing depth. The plot looked fine.
- **Check what PC1 correlates with** before interpreting it — library size, batch, RIN,
  collection date. One line each.
- **Look at the loadings.** "PC1 separates the groups" is an observation; "PC1 is the
  glucocorticoid response programme, led by FKBP5 and TSC22D3" is a result.
- **PC2 is not noise.** Here it is donor structure, which says the design should be
  `~ cell + dex`.
- **Reconstruction error is a quantitative alternative to the elbow** — where the data
  compresses well. It does not transfer to single-cell.
