---
title: "PCA on Microbiome Data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 5)
```

Microbiome data breaks one of PCA's assumptions in a way that bulk and single-cell
RNA-seq do not, and the failure is visible rather than subtle.

Relative abundances are **compositional**. Every sample sums to 1, so if one taxon goes
up another must come down — a negative correlation structure that exists because of the
constraint, not because of biology. The data lives on a simplex, not in the Euclidean
space PCA assumes.

This notebook shows the failure, the CLR fix, the alternative the field uses more often
(PCoA on Bray-Curtis), and the two tests that usually accompany it — plus the check that
usually gets skipped.

**Data:** `GlobalPatterns` — 16S sequencing across nine environment types, from soil and
ocean to human gut and skin. Caporaso et al. (2011), *PNAS* 108:4516.

```{r}
library(phyloseq)
library(compositions)
library(vegan)

pal <- c("#1B6B70", "#E07A5F", "#2A4B7C", "#81B29A", "#F2CC8F",
         "#8E7DBE", "#D62828", "#6A994E", "#BC4749")

data(GlobalPatterns)
GlobalPatterns
```

---

## Filtering

19,216 taxa across 26 samples, most near-zero. Keeping taxa with reasonable total
abundance and presence in more than five samples:

```{r}
gp <- prune_taxa(taxa_sums(GlobalPatterns) > 100, GlobalPatterns)
gp <- prune_taxa(rowSums(otu_table(gp) > 0) > 5, gp)
gp

env  <- sample_data(gp)$SampleType
cols <- pal[as.integer(factor(env))]
table(env)
```

Note the group sizes: two to four samples each. That matters later.

---

## PCA on relative abundances

The obvious thing to do, and what many pipelines still do.

```{r}
gp_rel <- transform_sample_counts(gp, function(x) x / sum(x))

X_rel <- t(as(otu_table(gp_rel), "matrix"))
p_rel <- prcomp(X_rel, center = TRUE, scale. = FALSE)

plot(p_rel$x[, 1], p_rel$x[, 2], col = cols, pch = 19,
     xlab = "PC1", ylab = "PC2", main = "PCA on relative abundance")
```

Three points far right, everything else crushed against the axis. Soil, ocean, gut,
skin and tongue — genuinely different ecosystems — are indistinguishable.

---

## Diagnosing it

```{r}
vp_rel <- round(100 * p_rel$sdev[1:5]^2 / sum(p_rel$sdev^2), 1)
vp_rel
```

PC1 takes 36.1% of the variance. Which taxa build that axis?

```{r}
top <- sort(abs(p_rel$rotation[, 1]), decreasing = TRUE)[1:5]
tax_table(gp)[names(top), c("Phylum", "Genus")]
```

Cyanobacteria, including *Dolichospermum*. And the samples at the extremes:

```{r}
data.frame(type = env, PC1 = round(p_rel$x[, 1], 3))[order(-p_rel$x[, 1]), ][1:5, ]
```

All three are freshwater creek. This is a cyanobacterial bloom.

One taxon at high relative abundance in three samples consumed the entire first
component. Because abundances are bounded and sum to 1, a dominant organism does not
merely add variance — it suppresses everything else in those samples at the same time.
PCA reads that as the strongest signal in the dataset, which numerically it is.

---

## The fix: centred log-ratio

CLR divides each count by the sample's geometric mean and takes the log, removing the
sum-to-one constraint and moving the data into Euclidean space. PCA on CLR-transformed
data is sometimes called Aitchison PCA.

The `+ 1` pseudocount handles zeros, since log(0) is undefined. Zero handling in
compositional data is a genuine open problem and more principled imputation methods
exist; for real work they are worth using.

```{r}
X_counts <- t(as(otu_table(gp), "matrix"))
X_clr <- compositions::clr(X_counts + 1)
p_clr <- prcomp(X_clr, center = TRUE, scale. = FALSE)

vp_clr <- round(100 * p_clr$sdev[1:5]^2 / sum(p_clr$sdev^2), 1)
vp_clr
```

```{r echo=FALSE}
dir.create("../figures", showWarnings = FALSE)
dir.create("../results", showWarnings = FALSE)
```

```{r fig.width=11, fig.height=5}
layout(matrix(c(1, 2, 3), nrow = 1), widths = c(1, 1, 0.5))
par(mar = c(4.5, 4.5, 3, 1))

plot(p_rel$x[, 1], p_rel$x[, 2], col = cols, pch = 19,
     xlab = paste0("PC1 (", vp_rel[1], "%)"),
     ylab = paste0("PC2 (", vp_rel[2], "%)"), main = "Relative abundance")

plot(p_clr$x[, 1], p_clr$x[, 2], col = cols, pch = 19,
     xlab = paste0("PC1 (", vp_clr[1], "%)"),
     ylab = paste0("PC2 (", vp_clr[2], "%)"), main = "CLR-transformed")

par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", levels(factor(env)), col = pal[1:nlevels(factor(env))],
       pch = 19, cex = 0.8, bty = "n")
```

```{r echo=FALSE, results='hide'}
png("../figures/microbiome_pca_comparison.png", width = 2600, height = 1100, res = 300)
layout(matrix(c(1, 2, 3), nrow = 1), widths = c(1, 1, 0.5))
par(mar = c(4.5, 4.5, 3, 1))
plot(p_rel$x[, 1], p_rel$x[, 2], col = cols, pch = 19,
     xlab = paste0("PC1 (", vp_rel[1], "%)"),
     ylab = paste0("PC2 (", vp_rel[2], "%)"), main = "Relative abundance")
plot(p_clr$x[, 1], p_clr$x[, 2], col = cols, pch = 19,
     xlab = paste0("PC1 (", vp_clr[1], "%)"),
     ylab = paste0("PC2 (", vp_clr[2], "%)"), main = "CLR-transformed")
par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", levels(factor(env)), col = pal[1:nlevels(factor(env))],
       pch = 19, cex = 0.8, bty = "n")
dev.off()
```

PC1 falls from **36.1%** to **22.8%**, and every environment type now occupies its own
region.

Variance-explained is often read as a quality score. Here the higher number was one
taxon's abundance in three samples; the lower number is ecological structure across nine
environments. A scree plot cannot tell you which of those you have.

---

## Bray-Curtis and PCoA

CLR + PCA is the compositionally principled route. The field more often uses **PCoA on
Bray-Curtis distance**, which takes a different approach: rather than transforming the
data so Euclidean distance becomes meaningful, it defines a distance measure suited to
abundance data and ordinates that.

Bray-Curtis needs no pseudocount, but gives no loadings — you cannot read taxa off the
axes the way you can with PCA.

```{r}
d_bray <- vegdist(X_rel, method = "bray")
pcoa   <- cmdscale(d_bray, k = 5, eig = TRUE)

vp_pcoa <- round(100 * pcoa$eig[1:5] / sum(pcoa$eig[pcoa$eig > 0]), 1)
vp_pcoa
```

```{r}
plot(pcoa$points[, 1], pcoa$points[, 2], col = cols, pch = 19,
     xlab = paste0("PCo1 (", vp_pcoa[1], "%)"),
     ylab = paste0("PCo2 (", vp_pcoa[2], "%)"),
     main = "PCoA on Bray-Curtis")
legend("topright", levels(factor(env)), col = pal[1:nlevels(factor(env))],
       pch = 19, cex = 0.6, bty = "n")
```

```{r echo=FALSE, results='hide'}
png("../figures/pcoa_bray.png", width = 2000, height = 1600, res = 300)
par(mar = c(4.5, 4.5, 3, 1))
plot(pcoa$points[, 1], pcoa$points[, 2], col = cols, pch = 19,
     xlab = paste0("PCo1 (", vp_pcoa[1], "%)"),
     ylab = paste0("PCo2 (", vp_pcoa[2], "%)"),
     main = "PCoA on Bray-Curtis")
legend("topright", levels(factor(env)), col = pal[1:nlevels(factor(env))],
       pch = 19, cex = 0.6, bty = "n")
dev.off()
```

This recovers structure from the *relative abundances* that plain PCA could not — the
distance measure is doing the work the CLR transform did in the previous section.

But note that it arranges the samples differently. In the CLR ordination soil sat at the
extreme of PC1; here Mock and Feces occupy the extremes and soil sits in the middle.

```{r}
representation <- data.frame(
  representation = c("Relative abundance", "CLR-transformed counts",
                     "Bray-Curtis distances"),
  method    = c("PCA", "PCA", "PCoA"),
  axis1_pct = c(vp_rel[1], vp_clr[1], vp_pcoa[1]),
  axis2_pct = c(vp_rel[2], vp_clr[2], vp_pcoa[2])
)
representation
```

```{r echo=FALSE}
write.csv(representation, "../results/ordination_by_representation.csv",
          row.names = FALSE)
```

Same 26 samples. The first axis carries 36.1%, 22.8% or 14.8% of the variation
depending only on how the data is represented. Nobody looking at one of these
ordinations alone would guess that.

---

## Testing group differences

An ordination shows structure; it does not test it. Two tests are standard.

**PERMANOVA** (`adonis2`) partitions the distance matrix by group and tests whether
between-group distances exceed what permutation would produce. It reports R², the
proportion of variation attributable to group — which matters as much as the p-value.

```{r}
adonis2(d_bray ~ env)
```

R² = 0.733, p = 0.001. Environment accounts for 73% of the variation in community
composition.

**ANOSIM** compares ranked between-group and within-group distances, returning R between
-1 and 1.

```{r}
anosim(d_bray, env)
```

R = 0.988 — near-perfect separation. Worth remembering what these samples are, though:
soil, ocean, human gut, skin. About as different as microbial communities get. This is
not what to expect from, say, treated versus untreated within one body site.

---

## The check that is usually skipped

Both tests assume comparable within-group dispersion. If one group is far more variable
than another, either can return significance from that spread rather than from any
difference in location — and the ordination will look like separation.

`betadisper` tests it directly:

```{r}
bd <- betadisper(d_bray, env)
anova(bd)
```

**p = 0.0013.** Dispersion does differ between groups, so the PERMANOVA above is
ambiguous: the groups may differ in location, in dispersion, or both, and the test
cannot separate them.

```{r fig.width=8, fig.height=5.5}
par(mar = c(9, 4.5, 3, 1))
boxplot(bd, las = 2, cex.axis = 0.7, col = pal, border = "grey30",
        main = "Distance to group centroid (Bray-Curtis)",
        xlab = "", ylab = "Distance")
```

```{r echo=FALSE, results='hide'}
png("../figures/betadisper_boxplot.png", width = 2000, height = 1500, res = 300)
par(mar = c(9, 4.5, 3, 1))
boxplot(bd, las = 2, cex.axis = 0.7, col = pal, border = "grey30",
        main = "Distance to group centroid (Bray-Curtis)",
        xlab = "", ylab = "Distance")
dev.off()
par(mar = c(5, 4, 4, 2) + 0.1)
```

The spread is not subtle. Mock communities sit around 0.04 from their centroid, human
faeces around 0.44 — a tenfold difference. That is real biology, not an artifact: a
defined mock community is by construction more uniform than gut samples from different
people. But it is exactly the situation in which PERMANOVA becomes hard to interpret.

Freshwater and Tongue render as flat lines because they have only two samples each.

---

## Does the distance measure change the conclusion?

Euclidean distance on CLR-transformed data is **Aitchison distance** — a
compositionally-aware alternative to Bray-Curtis. Running the same two tests on it:

```{r}
d_aitch <- dist(X_clr)
adonis2(d_aitch ~ env)
```

```{r}
anova(betadisper(d_aitch, env))
```

```{r}
tests <- data.frame(
  distance              = c("Bray-Curtis", "Aitchison (CLR)"),
  permanova_R2          = c(0.733, 0.805),
  permanova_p           = c(0.001, 0.001),
  betadisper_p          = c(0.0013, 0.1006),
  dispersion_confounded = c("yes", "no")
)
tests
```

```{r echo=FALSE}
write.csv(tests, "../results/ordination_tests.csv", row.names = FALSE)
```

The compositionally-aware distance is better on both counts: a **larger** effect size
(0.805 against 0.733) and **no dispersion confounding** (p = 0.10 against p = 0.0013).

That is the strongest argument in this notebook for taking compositionality seriously.
CLR is not a cosmetic choice that makes the plot look tidier — on this data it changes
what the statistics are entitled to conclude. The Bray-Curtis result is significant but
ambiguous; the Aitchison result is significant and clean.

---

## What to take from this

- **Relative abundances are compositional.** Standard PCA assumes Euclidean geometry;
  compositional data does not live there.
- **A dominant taxon can consume PC1** without telling you anything about your question.
- **The ordination depends on the representation.** 36.1%, 22.8% or 14.8% on the first
  axis, from the same 26 samples.
- **Two routes work:** CLR + PCA keeps loadings so you can identify taxa; PCoA on
  Bray-Curtis handles zeros without a pseudocount but has no loadings. UniFrac adds
  phylogenetic distance, which neither does.
- **Ordination shows, it does not test.** And a test without `betadisper` and an effect
  size is hard to interpret.
- **Here the compositional route won on the statistics too**, not just the plot.
